---
title: "Run cuspra on empirical data"
author: "Romain Frelat"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Run cuspra on empirical data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


These are short explanations on how to use CUSPRA on empirical dataset, as introduced in Sguotti et al. (2024) "Resilience assessment in complex natural systems" *Proc. R. Soc. B.* 291:20240089 doi [10.1098/rspb.2024.0089](http://doi.org/10.1098/rspb.2024.0089)


## 1. Load package and dataset

Load the `cusp` and `cuspra` R packages. To make sure that you have the last version of the package, you can run `devtools::install_github(repo = "rfrelat/cuspra", dependencies = TRUE, build_vignettes = TRUE)`.

```{r load}
library(cusp)
library(cuspra)
```

Let's now load the dataset:

```{r data}
data("ecosystem_ns")
# or 
# data("fishtraits_med")
# data("cod_nea") 
```

If you use another dataset, make sure to change the name of the object in the cusp() model with the correct name of columns (e.g. `PC1` should be replaced by `Biomass` if you are using the `cod_nea` dataset).

## 2. Fit the cusp model

```{r cusp}
fitNS <- cusp(y ~ PC1, alpha ~ Fishing,
             beta ~ Temperature,  data=ecosystem_ns)
# Make sure to check the model results
# with summary(), evalcusp(), or visually with plot()
evalcusp(fitNS)
```

Make sure to check the model results with `summary(fitNS)`, `plot(fitNS)` and/or `evalcusp(fitNS)`. 
The output of `evalcusp()` shows:
- `rsquared`: the pseudo R2 of the cusp model (**should be >0.3**)}
- `delta.aicc`: the difference between AICc of the linear and cusp model (**should be >0**)}
- `percin`: percentage of points within the cusp area (**should be >10**)}
- `pval.state`: p-value of the state variable (**should be <0.05**)}


## 3. Estimate the resilience
```{r cuspra}
raNS <- cuspra(fitNS)
```

The output of the cusp resilience assessment provide 4 columns:
- cuspRA: the resilience assessment, with values between 0 (low resilience) and 1 (high resilience)
- dalpha: the horizontal component as distance to the cusp area (if negative, the point is inside the cusp area)
- dbeta: the vertical component as distance to linearity (if negative the system is discontinuous)
- sumd: the weighted average of the horizontal and vertical components (by default the horizontal component has double weight)

## 4. Visualize the results

```{r, fig.show='hold'}
layout(matrix(c(2,1,1), ncol=1))
pal <- plotra(raNS$cuspRA, fitNS)
par(mar=c(2,4,1,1))
plot(ecosystem_ns$Year, ecosystem_ns$PC1, pch=16,
     ylab="State variable", type="b",
     col=as.character(cut(raNS$cuspRA, breaks = pal$bk, labels = pal$pal)))
```
